% Copyright (C) 2014-2016 by Thomas Auzinger <thomas@auzinger.name>

\documentclass[final]{vutinfth} % Remove option 'final' to obtain debug information.

% Load packages to allow in- and output of non-ASCII characters.
\usepackage{lmodern}        % Use an extension of the original Computer Modern font to minimize the use of bitmapped letters.
\usepackage[T1]{fontenc}    % Determines font encoding of the output. Font packages have to be included before this line.
\usepackage[utf8]{inputenc} % Determines encoding of the input. All input files have to use UTF8 encoding.

\usepackage{xfrac}

% Extended LaTeX functionality is enables by including packages with \usepackage{...}.
\usepackage{fixltx2e}   % Provides fixes for several errors in LaTeX2e.
\usepackage{amsmath}    % Extended typesetting of mathematical expression.
\usepackage{amssymb}    % Provides a multitude of mathematical symbols.
\usepackage{mathtools}  % Further extensions of mathematical typesetting.
\usepackage{microtype}  % Small-scale typographic enhancements.
\usepackage{enumitem}   % User control over the layout of lists (itemize, enumerate, description).
\usepackage{multirow}   % Allows table elements to span several rows.
\usepackage{booktabs}   % Improves the typesettings of tables.
\usepackage{subcaption} % Allows the use of subfigures and enables their referencing.
\usepackage[ruled,linesnumbered,algochapter]{algorithm2e} % Enables the writing of pseudo code.
\usepackage[usenames,dvipsnames,table]{xcolor} % Allows the definition and use of colors. This package has to be included before tikz.
\usepackage{nag}       % Issues warnings when best practices in writing LaTeX documents are violated.
\usepackage{hyperref}  % Enables cross linking in the electronic document version. This package has to be included second to last.
\usepackage[acronym,toc]{glossaries} % Enables the generation of glossaries and lists fo acronyms. This package has to be included last.

\usepackage[final]{pdfpages}
\usepackage{epstopdf}

\usepackage{appendix}
\renewcommand\appendixname{New Name}

% Define convenience functions to use the author name and the thesis title in the PDF document properties.
\newcommand{\authorname}{Mathieu Le Muzic} % The author name without titles.
\newcommand{\thesistitle}{{From Atoms to Cells: Interactive and Illustrative Visualization of Digitally Reproduced Lifeforms}} % The title of the thesis. The English version should be used, if it exists.

% Set PDF document properties
\hypersetup{
    pdfpagelayout   = TwoPageRight,           % How the document is shown in PDF viewers (optional).
    linkbordercolor = {Melon},                % The color of the borders of boxes around crosslinks (optional).
    pdfauthor       = {\authorname},          % The author's name in the document properties (optional).
    pdftitle        = {\thesistitle},         % The document's title in the document properties (optional).
    pdfsubject      = {Subject},              % The document's subject in the document properties (optional).
    pdfkeywords     = {a, list, of, keywords} % The document's keywords in the document properties (optional).
}

\setsecnumdepth{subsection} % Enumerate subsections.

\nonzeroparskip             % Create space between paragraphs (optional).
\setlength{\parindent}{0pt} % Remove paragraph identation (optional).

\makeindex      % Use an optional index.
\makeglossaries % Use an optional glossary.
%\glstocfalse   % Remove the glossaries from the table of contents.

% Set persons with 4 arguments:
%  {title before name}{name}{title after name}{gender}
%  where both titles are optional (i.e. can be given as empty brackets {}).
\setauthor{}{\authorname}{}{male}
\setadvisor{Dr.techn}{Ivan Viola}{}{male}

% For bachelor and master theses:
\setfirstassistant{Pretitle}{Forename Surname}{Posttitle}{male}
\setsecondassistant{Pretitle}{Forename Surname}{Posttitle}{male}
\setthirdassistant{Pretitle}{Forename Surname}{Posttitle}{male}

% For dissertations:
\setfirstreviewer{Pretitle}{Forename Surname}{Posttitle}{male}
\setsecondreviewer{Pretitle}{Forename Surname}{Posttitle}{male}

% For dissertations at the PhD School:
\setsecondadvisor{Pretitle}{Forename Surname}{Posttitle}{male}

% Required data.
\setaddress{Apartment 15, The Needleworks 41-43, Albion Street, Leicester LE1 6GF, United Kingdom}
\setregnumber{1326132}
\setdate{03}{10}{2016} % Set date with 3 arguments: {day}{month}{year}.
\settitle{\thesistitle}{From Atoms to Cells: Interactive and Illustrative Visualization of Digitally Reproduced Lifeforms}


 % Sets English and German version of the title (both can be English or German).
%\setsubtitle{Optional Subtitle of the Thesis}{Optionaler Untertitel der Arbeit} % Sets English and German version of the subtitle (both can be English or German).

% Select the thesis type: bachelor / master / doctor / phd-school.
% Bachelor:
%\setthesis{bachelor}
%
% Master:
%\setthesis{master}
%\setmasterdegree{dipl.} % dipl. / rer.nat. / rer.soc.oec. / master
%
% Doctor:
\setthesis{doctor}
\setdoctordegree{techn.}% rer.nat. / techn. / rer.soc.oec.
%
% Doctor at the PhD School
%\setthesis{phd-school} % Deactivate non-English title pages (see below)

% For bachelor and master:
\setcurriculum{Media Informatics and Visual Computing}{Medieninformatik und Visual Computing} % Sets the English and German name of the curriculum.

% For dissertations at the PhD School:
\setfirstreviewerdata{Affiliation, Country}
\setsecondreviewerdata{Affiliation, Country}

% Define convenience macros.
\newcommand{\todo}[1]{{\color{red}\textbf{TODO: {#1}}}} % Comment for the final version, to raise errors.


\begin{document}

\frontmatter % Switches to roman numbering.
% The structure of the thesis has to conform to
%  http://www.informatik.tuwien.ac.at/dekanat

\addtitlepage{naustrian} % German title page (not for dissertations at the PhD School).
\addtitlepage{english} % English title page.
\addstatementpage

%\begin{danksagung*}
%\todo{Ihr Text hier.}
%\end{danksagung*}

%\begin{acknowledgements*}
%\todo{Enter your text here.}
%\end{acknowledgements*}

\begin{abstract}
	
	Macromolecules, such as proteins, are the building blocks of the machinery of life, and therefore are essential to the comprehension of physiological processes.
	In physiology, illustrations and animations are often utilized as a mean of communication because they can easily be understood with little background knowledge.
	However, their realization requires numerous months of manual work, which is both expensive and time consuming.	
	Computational biology experts produce everyday large amount of data that is publicly available and that contains valuable information about the structure and also the function of these macromolecules.	
	Instead of relying on manual work to generate illustrative visualizations of the cell biology, we envision a solution that would utilize all the data already available in order to streamline the creation process.

	In this thesis are presented several contributions that aim at enabling our vision.
	First, a novel GPU-based rendering pipeline that allows interactive visualization of realistic molecular datasets comprising up to hundreds of millions of macromolecules.
	The rendering pipeline is embedded into a popular game engine and well known computer  graphics optimizations were adapted to support this type of data, such as level-of-detail, instancing and occlusion queries.
	Secondly, a new method for authoring cutaway views and improving spatial exploration of crowded molecular landscapes. 
	The system relies on the use of clipping objects that are manually placed in the scene and on visibility equalizers that allows fine tuning of the visibility of each species present in the scene.	
	Agent-based modeling produces trajectory data that can also be combined with structural information in order to animate these landscapes.
	The snapshots of the trajectories are often played in fast-forward to shorten the length of the visualized sequences, which also renders potentially interesting events occurring at a higher temporal resolution invisible.
	The third contribution is a solution to visualize time-lapse of agent-based simulations that also reveals hidden information that is only observable at higher temporal resolutions.
	And finally, a new type of particle-system that utilize quantitative models as input and generate missing spatial information to enable the visualization of molecular trajectories and interactions. 
	The particle-system produces a similar visual output as traditional agent-based modeling tools for a much lower computational footprint and allows interactive changing of the simulation parameters, which was not achievable with previous methods.
%	These new visualization technologies are showcased with accurate scientific data which was collected with the help of experts, and were also designed and validated in collaboration with renowned scientific illustrators.
		
\end{abstract}

\begin{kurzfassung}
	Makromoleküle, wie z.B. Proteine, sind die grundlegenden Bausteine von der Maschinen des Lebens und somit essentiell für das Verständnis von physiologischen Prozessen.
	Illustrationen und Animationen dienen häufig als Mittel für die Kommunikation in der Physiologie, weil sie einfach und mit wenig Hintergrundwissen zu verstehen sind.
	Ihre Entwicklung braucht jedoch oft Monate von teurer und zeitintensiver manueller Arbeit.
	Experten aus dem Bereich von Computational Biology produzieren jeden Tag riesige Mengen von öffentlich zugänglichen Daten, welche wertvolle Informationen über die Strukturen und Funktionen von Makromoleküle enthalten.
	Wir haben die Vision den Erstellungsprozess von illustrativen Visualisierungen im Bereich der Zellbiologie zu optimieren, wobei die gesamten verfügbaren Daten benutzt werden, anstatt auf rein manuelle Arbeit angewiesen zu sein.
	
	Im Rahmen dieser Arbeit werden mehrere Ansätze vorgestellt, welche darauf abzielen diese Vision zu realisieren.
	Der erste Teil dieser Arbeit beschreibt eine Rendering Pipeline, welche interaktive Visualisierung von realistischen molekularen Datensätze ermöglicht mit Hunderten Millionen Markromolekülen.
	Die Rendering Pipeline ist in einer weit verbreiteten Spiele Engine entwickelt und bekannte Computergraphik Optimierungen wie level-of-detail, instancing und occlusion queries wurden adaptiert wurden um die molekularen Datensätze zu unterstützen.
	Im zweiten Teil dieser Arbeit wird eine neue Methode für die Erstellung von Cutaway Views präsentiert, welche die räumlichen Erkundung von überfüllten molekularen Landschaften verbessert.
	Die Methode basiert auf der Benutzung von Clipping-Objekten, welche manuell in der Szene platziert werden und einem Visibility-Equalizer, welcher den visuellen Feinabgleich von den verschiedenen Objekten in der Szene ermöglicht.
	Um solche Szene zu animieren wird häufig agenten-basierte Modellierung eingesetzt, welche Raumkurven produziert die mit mit strukturellen Information kombiniert werden können.
	Momentaufnahmen von den Raumkurven werden häufig im Schnellablauf abgespielt um die Länge der visualisierten Sequenz zu verkürzen, wodurch jedoch auch potentiell interessante Ereignisse nicht mehr sichtbar sein können.
	Im dritten Teil dieser Arbeit wird ein Ansatz präsentiert, der Zeitrafferaufnahmen von agentenbasierten Simulationen visualisiert, welcher versteckte Informationen aufdeckt, die nur mit eine höheren zeitlichen Auflösung sichtbar wären.
	Der letzte Teil dieser Arbeit beschreibt einen neuen Typ von Partikelsystemen, welcher quantitative Modelle als Input benutzt und fehlender räumliche Information generiert, um die Visualisierungen von Molekularen Raumkurven und Interaktionen zu visualisieren.
	Das Partikelsystem produziert einen ähnlichen visuellen Output, wie traditionelle agentenbasierten Modelle, allerdings mit geringeren Rechenaufwand.
	Zusätzlich erlaubt das System interaktiv Simulationsparameter zu verändern, was mit bisherigen Methoden nicht möglich gewesen ist.
	
\end{kurzfassung}



% Select the language of the thesis, e.g., english or naustrian.
\selectlanguage{english}

% Add a table of contents (toc).
\tableofcontents % Starred version, i.e., \tableofcontents*, removes the self-entry.

% Switch to arabic numbering and start the enumeration of chapters in the table of content.
\mainmatter

\input{body.tex} % The body of the thesis




\backmatter

% Use an optional list of figures.
%\listoffigures % Starred version, i.e., \listoffigures*, removes the toc entry.

% Use an optional list of tables.
%\listoftables % Starred version, i.e., \listoftables*, removes the toc entry.

% Use an optional list of alogrithms.
%\listofalgorithms
%\addcontentsline{toc}{chapter}{List of Algorithms}

% Add an index.
\printindex

% Add a glossary.
%\printglossaries

% Add a bibliography.
\bibliographystyle{alpha}
\bibliography{bibliography}

\end{document}